Time for a rethink: time sub-sampling methods in disparity-through-time analyses

### 2018-02-07 manuscript/response_to_reviewers.tex

We apologise for the confusion in the description of Figure 3 and have fixed it (see Reviewer 1, comment \ref{fig3_feckup}). We were rushing to get this submitted by the symposium deadline and this error slipped through the cracks.

\item{\textcolor{blue}{Finally, I have spent some time thinking about by the statement ``all time binning approaches (whether equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model.'' (lines 70-71). I see what the authors are getting at — they are worried that doing this requires an assumption of constant evolution across the time bin. But I would argue that it is not necessary to interpret it this way. Instead it could simply represent an average across the interval sampled without the need to make any assumption about how change occurred within that interval. Plus, even in scenarios where there was enough change that at a course scale, it appeared that disparity was static, some of that volatility might be revealed if the time-scale were altered.}}

We've toned down this statement by including the fact that bins can simply represent averages (as suggested by the reviewer).

\textit{Second, the time binning approaches (whether bins are equally sized or not) favour punctuated equilibrium modes of evolution.
Whether the disparity represents an average across the interval (with no interpretation of if or how it varied within the time bin), or it is effectively postulated to be constant, when analysing the changes in disparity-through-time, this method will only allow changes in disparity to occur between intervals rather than also allowing for gradual changes within intervals (a pattern that is fairly common in the fossil record, CITATION).} lines @@@

%TG: tone done the statement? % NC: yeah, not sure what else we can do. %TG: Does that kinda work? NC: sounds ok to me, I tidied the english.

@@

\item{\textcolor{blue}{But there is also a deeper issue here, in that the fossil record is at its heart a time-binned record: rates of sedimentation vary and sediment deposition is sporadic, such that local beds represent different amounts of time, or different ``packages'' of time, separated by gaps, and the fossils within them may be time-averaged to various degrees. [...] Therefore, even if you pick a ``point in time'' and count the number of branches/ranges that cross that point, some of those are being captured because the resolution of the end points only allows us to know that the taxon started and/or ended somewhere in the interval encompassing that point in time.}}

We've added the following caveat to the discussion:

\textit{Finally, the difference between the time-binning and time-slicing results could also simply be due to the nature of the fossil record.
Rates of sedimentation vary in time and space influencing the groups found within the rock record and their temporal distribution. 
In this case, different beds could represent different ``packages'' of fauna through time separated by gaps, resulting in natural ``bins'' rather than slices of the data.
Slicing through such strata will yield similar results no matter where in time the slice occurs.
It is important to note however, that the time slicing method also includes ancestral estimations (either through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.
Additionally, this effect is likely to be most obvious in vertebrates and less problematic for groups like Foraminifera.} lines @@@

% NC: Had to edit this quite a bit to make the English make sense. Does it still convey your meaning wrt forams?

@@

\subsection{Reviewer 2 - Minor comments}

\begin{enumerate}

\item{\textcolor{blue}{It is not immediately clear what is meant by ``defined by the average duration of the stratigraphic period'' (line 337, 341) vs ``defined by the average number of stratigraphic periods'' (line 348, 353). After thinking about it for awhile and then checking the R code, this is what appears to be going on.
Let's say we have a case study that spans 10 stratigraphic intervals over 40 million years.
Thus for the stratigraphy approach, there are 10 sub-samples which vary in their duration.
Let’s say the average duration of those intervals is 5 million years.
For the duration approach, the 40 million year span is divided into 5-million-year bins, resulting in 8 sub-samples.
For the number approach, the 40 million year span is divided into 10, each of which is 4 million years long.
If this is the case, the use of the word ``average'' when describing the number approach is confusing. I am also having trouble reconciling this with the number of points in each panel in Figure 2, unless there is also a minimum sample size applied (for example in the Bapst 2016 study, there are only three disparity values for the time-binned series using the ``number'' approach, maybe because after dividing the time scale up this way, there were not enough taxa in the first and second time bins?).
It may be helpful to include a graphic, even if in relegated to one of the appendices.}}

We've added a figure describing the different time-sampling strategies and changed the relevant methods section to be clearer. % NC: Haven't changed the methods yet.

### 2018-02-12 manuscript/response_to_reviewers.tex

\begin{enumerate}

\item{\textcolor{blue}{In the section on adding in reconstructed ancestors to the dataset before ordination, I suppose you should mention that this is similar to the approach that we outlined in our Paleobiology paper (Brusatte et al. 2011) and mention any key differences.}}

We now explicitly refer to Brusatte et al. 2011 and mention the key difference:

\textit{This approach is similar to Brusatte et al. (2011) but uses model based estimations (rather than parsimony) allowing to control for ambiguous nodes (i.e. poorly estimated).} lines 141-144.
%TG: checked and matched with manuscript.

@@

\item{\textcolor{blue}{I'm not sure about the use of the term ‘cladisto-spaces’. To me this implies that the space itself is ordinated using phylogenetic information (e.g., branch lengths). Now, you are indeed including some phylogenetic information in your morphospaces, in terms of reconstructed ancestors and inferred states for branches. So sure, I think what you have could be called a ‘cladisto-space’. Same with most of the morphospace options produced by Claddis. But the ones that most of us have done in the past, simply using discrete characters to ordinate a morphospace, haven't generally incorporated phylogeny. I’m talking about the references you cite in lines 141-143 here. These morphospaces been ordinated using discrete characters of the type that are used in phylogenetic analysis. But the characters are not phylogenetically transformed in any way. As far as the ordination analysis knows, they are simply discrete character statements. So I wonder if the term ‘cladisto-space’ is confusing, and is just yet another new term that may muddy the waters rather than clarify. If you want to keep using this term, that's fine, but a sentence or two of explanation (about how previous authors have used discrete characters in a non-phylogenetic way) would be warranted.}}

We've change the term cladisto-space to morphospace throughout the manuscript.
Although we prefer the ``cladisto-space'' term, we agree with this concern about adding a ``new term that may muddy the waters rather than clarify''.
We've emphasised the differences, however, with other uses of the word ``morphospace'' in an explanatory paragraph:

\textit{To explore disparity-through-time in our datasets, we used a morphospace approach (CITATIONS).
Morphospaces can be obtained from any multidimensional morphological data set but can differ in the data used (e.g. discrete or continuous), and whether they include phylogenetic data or not. 
Although empirical morphospaces from discrete or continuous data have been shown to have similar properties (CITATIONS), our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis;} c.f. \textit{geometric morphometric data) and includes some phylogenetic information (see above).
Mathematically, our morphospaces are $n$ dimensional objects that summarise the distances between discrete morphological characters of the taxa present and their ancestors.} lines 145-158
%TG: checked and matched with manuscript.

@@

We have done as suggested - we didn't do this the first time round because of issues with ggplot. We also made the same change in the appendix figures, and changed the colours in response to a comment by reviewer 2.

\item{\textcolor{blue}{I may be misinterpreting Figure 3, but it seems like peak disparity values for the Brusatte et al. dataset are pretty much the same regardless of method used (second plot from top). This doesn't jibe with the statement in the text (lines 435-437). Please clarify.}}
\label{fig3_feckup}

We apologise for mismatching the datasets and the patterns described in the text.
We've now corrected this error and double check this as follows:

\textit{In the Beck \& Lee (2014) and Bapst et al. (2016) datasets, disparity peaks occur much at much older ages when time-slicing rather than time binning approaches are used (Fig. 3; Supporting Information Appendix S2: Figs A3-A4).
This is also true for stratigraphic time bins in the Wright (2017) dataset, although when using equal time bins the peaks are later than the time-slicing methods, or very similar (Fig. 3; Supporting Information Appendix S2: Figs A3-A4).
Across the three time binning methods, the Brusatte et al. (2014) dataset has similar disparity peaks whichever method is used, the Wright (2017) dataset only had variation in peaks when using unequal time bins (stratigraphy), whereas in the Bapst et al. (2016) and Beck \& Lee (2014) datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Fig. 3; Supporting Information Appendix S2: Figs A3-A4).} lines 448-461
%TG: checked and matched with manuscript.

@@

We've also added a note about discrepancies in the time-slicing models in Beck \& Lee (2014):

\textit{Additionally, there seems to be small discrepancies within the time-slicing methods (gradual vs. proximity) except in the Beck \& Lee (2014) dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig. 3; Supporting Information Appendix S2: Figs A3-A4).} lines 461-465
%TG: checked and matched with manuscript.

@@

\begin{enumerate}

\item{\textcolor{blue}{First, if I understand lines 278-299 (and the R annotation) correctly, the gradual evolution models are not in fact fundamentally different from the punctuated models. They are both assigning scores along the branch that are the same as either the ancestor or descendant — the difference is only that the selection is based on a probability in the ``gradual'' models, rather than a rule, as in the ``punctuated'' models. The concept of gradual change would be more closely modelled by assigning values along a branch that were in-between those assigned to the ancestor and descendent but based on how far along the branch any given point is. For example, let's say the value of the ancestor along PCO1 is 0.04 and that of the descendent is 0.03. For a point that is halfway along the branch linking them, the value should be 0.035 (instead of assigning either 0.03 or 0.04 based on a probability). This, of course, makes assumptions about the course of evolution between the ancestor and descendant, but at least would be describing gradual change.}}

The reviewer does indeed understand the \texttt{equal.splits} and \texttt{gradual.splits} models correctly: they select the values in the matrix from the edge's descendant or ancestor with a probability value for each.
We believe however that the reviewer's proposition of making these models ``actually'' gradual (e.g. by selecting the mean value between two points) might lead to incorrect interpretation of the results.

The rational for proposing the \texttt{equal.splits} and \texttt{gradual.splits} models rather than a pure gradual model is to not create any extra data points in the ordinated space.
In fact, doing so pre-ordination can change the ordinated space geometry, while doing so post-ordination leads to the incorrect estimation of the space: (1) the number of elements changes hence changing parameters for future statistical tests (e.g. degrees of freedom) and (2) the variance per axis will change as well.
Our methods allow us to approach a pure gradual model without modifying the ordinated space's structure and properties.
In fact, when applying the \texttt{equal.split} model (or \texttt{gradual.splits}) on a decent number of bootstrap replicates (say 100), the ancestor's PCO1 value (0.04 in the reviewer's example) will be sampled 50\% of the time and the same for the descendant's value (0.03 - with the probability changing as a function of branch length for the \texttt{gradual.splits} model).
The resulting metric would then be based on the following PCO1 value: $0.5\times0.04 + 0.5\times0.03 = 0.035$ thus identical to what the reviewer suggests, without modifying the multidimensional space.

@@

We realise we had not made this clear, thus we have added the following explanation in the text:

\textit{For example, using the ``equal.splits'' model on an ancestor and a descendant with PCO1 values of respectively $0.04$ and $0.03$, after a sufficient number of bootstrap replicates (e.g. 100) the value along the branch will be close to $0.5\times0.04 + 0.5\times0.03 = 0.035$.
By estimating this value rather than generating it (i.e. creating a new element mid-way along the branch that would be the average of the descendant and ancestor - $0.035$) we obtain the same results without modifying the morphospace properties.} lines 318-325
%TG: checked and matched with manuscript.

@@

%TG: tone done the statement? % NC: yeah, not sure what else we can do. %TG: Does that kinda work? NC: sounds ok to me, I tidied the english.

\item{\textcolor{blue}{But there is also a deeper issue here, in that the fossil record is at its heart a time-binned record: rates of sedimentation vary and sediment deposition is sporadic, such that local beds represent different amounts of time, or different ``packages'' of time, separated by gaps, and the fossils within them may be time-averaged to various degrees. [...] Therefore, even if you pick a ``point in time'' and count the number of branches/ranges that cross that point, some of those are being captured because the resolution of the end points only allows us to know that the taxon started and/or ended somewhere in the interval encompassing that point in time.}}

We've added the following caveat to the discussion:

\textit{The difference between the time-binning and time-slicing results could also simply be due to the nature of the fossil record.
Rates of sedimentation vary in time and space influencing the groups found within the rock record and their temporal distribution. 
In this case, different beds could represent different ``packages'' of fauna through time separated by gaps, resulting in natural ``bins'' rather than slices of the data.
Slicing through such strata will yield similar results no matter where in time the slice occurs.
It is important to note however, that the time slicing method also includes ancestral estimations (either through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.
Additionally, this effect is likely to be most obvious in vertebrates or groups where the fossil is ``patchy'' but less problematic for groups like Foraminifera.} lines @@@

@@

% NC: Had to edit this quite a bit to make the English make sense. Does it still convey your meaning wrt forams?

\item{\textcolor{blue}{As an aside, the Hunt et al 2015 citation does not clarify the matter (line 74-75). A case could be made that all time series are binned (if you consider that the samples are likely time-averaged and there are gaps between samples--although this is different from dividing up "everything" into adjacent boxes), and yet there are time series that are best characterized by gradual directional change (e.g., best fit by a biased random walk). In addition, Hunt et al found that directional change was actually more likely to be detected in series of long duration. Furthermore, homogenous rate patterns were more likely to be supported than heterogenous/punctuated change with time series of fewer samples, which includes cases where each ``step'' includes a longer interval of time over which shorter changes could be hidden.}}

%TG: not sure where the reviewer want us to go with that. Is that the same nuancing point as the comment before? In that case we can simply merge them and make a caveat paragraph. 
% NC: Yeah I'm not sure either, but maybe needs a paragraph in the discussion? Currently the discussion is short (and written by me in about an hour!) so would be good to add something there.

\textit{Finally, Hunt et al 2015 found that time series are best characterized by gradual directional changes (biased random walks).
In fact, homogeneous directional changes are more likely to be supported than heterogeneous ones (e.g. punctuated changes) in longer duration series with few samples per series.
In our implementation of time-slicing, the models are not selected based on any fit value (e.g. AIC) but merely on researcher's assumptions.
We thus suggest that both types of models (punctuated and gradual) are tested during analysis unless there could be strong independent support for one or the other.}

### 2018-02-12 manuscript/time-slice.tex
Added last comment from reviewer 2 + double checked reviewer 1's

\paragraph{Building morphospaces.} 
To explore disparity-through-time in our datasets, we used a morphospace approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
Morphospace can be obtained from any multidimensional morphological data set but can differ in the data used (e.g. discrete or continuous) or whether they include phylogenetic data or not.
Although empirical morphospaces from discrete or continuous data have been shown to have similar properties \citep{foth2012different,hetherington2015cladistic}, we would like to highlight that our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis; \textit{c.f.} geometric morphometric data) and includes some phylogenetic information (see above).
Mathematically, our morphospaces are $n$ dimensional objects that summarises the distances between discrete morphological characters of the taxa present and their ancestors.

@@

It is important to note that the time-slicing method is not an ancestral states estimation method \textit{per se}.
This method does not estimate values along a branch applying a model (\textit{c.f.} methods described for ancestral character estimation in the ``Preparing the data for disparity-through-time analysis'' section above) but rather chooses between the two available pieces of information (the ordination score of the descendant or the ancestor) using the methods described above.
This allows the method to be used in post-ordination analysis where the data used in each time-slice is data already present in the morphospace.
In other words, this method does not require a re-ordination of the morphospace every time a slice goes through a branch, thus allowing the properties of the morphospace (e.g. distance between species, variance of each axis, etc.) to remain constant.
For example, using the ``equal.split'' model between an ancestor and a descendant with PCO1 values of respectively $0.04$ and $0.03$, after a sufficient number of bootstrap replicates (e.g. 100) the value along the branch will be close to $0.5\times0.04 + 0.5\times0.03 = 0.035$.
By estimating this value rather than generating it (i.e. creating a new element midway along the branch that would be the average of the descendant and ancestor - $0.035$) we obtain the same results without modifying the morphospace properties.

@@

\paragraph{Disparity peaks.}
In the \cite{beckancient2014} and \cite{bapst2016topology} datasets, disparity peaks occur much at much older ages when time-slicing rather than time binning approaches are used (Fig. \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
This is also true for stratigraphic time bins in the \cite{wright2017bayesian} dataset, although when using equal time bins the peaks are later than the time-slicing methods, or very similar (Fig. \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
Across the three time binning methods, the \cite{brusatte2014gradual} dataset have similar disparity peaks whichever method is used, the \cite{wright2017bayesian} dataset only had variation in peaks when using un-equal time bins (stratigraphy) whereas in the \cite{bapst2016topology} and \cite{beckancient2014} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Fig. \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
Additionally, there seem to be low discrepancy within the time-slicing methods (gradual \textit{vs.} proximity) expect in \cite{beckancient2014} dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig. \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).

@@

Of course the method also has limitations.
The main one of these is a practical one; it requires a time-calibrated phylogeny which are not available for all palaeontological datasets.
Furthermore, like most phylogeny based methods, time-slicing depends on ancestral state estimations.
Care should be taken in interpreting these, as they are highly dependent on the data and models used for the estimations \citep{Slateretal2012,doi:10.1080/10635150801910451}.
Finally, the difference between the time-binning and time-slicing methods could also simply be favoured towards time binning due to the nature of the fossil record itself.
The rates of sedimentation varies in time a space and can be really sporadic, influencing the groups found within and their temporal distribution (e.g. vertebrates, c.f. foraminifera).
In this case, different beds could represent different ``packages'' of fauna through time separated by gaps thus effectively binning the data rather than slicing.
For example, slicing through such strata will yield similar results whether where in time does the slice occur.
It is important to note however, that, as specified above, the time slicing method also includes ancestral estimations data (whether through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.

@@

\paragraph{Acknowledgments.}
  NC thanks Mark Sutton and Philip Mannion for the invitation to contribute to the `Evolutionary Modelling' symposium at The Palaeontological Association Annual Meeting 2017.
  TG acknowledges support from the Australian Discovery Project Grant number DP170103227 awarded to Vera Weisbecker.
  We thank Dave Bapst, Graeme Lloyd, April Wright and Davey Wright for assistance in gathering data for the analyses and/or discussions about the approach; and Steve Brusatte and one anonymous reviewer for helpful comments on the manuscript.

>>>>

\paragraph{Building morphospaces.} 
To explore disparity-through-time in our datasets, we used a morphospace approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
Morphospaces can be obtained from any multidimensional morphological data set but can differ in the data used (e.g. discrete or continuous), and whether they include phylogenetic data or not.
Although empirical morphospaces from discrete or continuous data have been shown to have similar properties \citep{foth2012different,hetherington2015cladistic}, our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis; \textit{c.f.} geometric morphometric data) and includes some phylogenetic information (see above).
Mathematically, our morphospaces are $n$ dimensional objects that summarise the distances between discrete morphological characters of the taxa present and their ancestors.

@@

It is important to note that the time-slicing method is not an ancestral states estimation method \textit{per se}.
This method does not estimate values along a branch applying a model (\textit{c.f.} methods described for ancestral character estimation in the ``Preparing the data for disparity-through-time analysis'' section above) but rather chooses between the two available pieces of information (the ordination score of the descendant or the ancestor) using the methods described above.
This allows the method to be used in post-ordination analysis where the data used in each time-slice is data already present in the morphospace.
In other words, this method does not require a re-ordination of the morphospace every time a slice goes through a branch, thus allowing the properties of the morphospace (e.g. distance between species, variance of each axis, etc.) to remain constant.
For example, using the ``equal.splits'' model on an ancestor and a descendant with PCO1 values of respectively $0.04$ and $0.03$, after a sufficient number of bootstrap replicates (e.g. 100) the value along the branch will be close to $0.5\times0.04 + 0.5\times0.03 = 0.035$.
By estimating this value rather than generating it (i.e. creating a new element mid-way along the branch that would be the average of the descendant and ancestor - $0.035$) we obtain the same results without modifying the morphospace properties.

@@

\paragraph{Disparity peaks.}
In the \cite{beckancient2014} and \cite{bapst2016topology}) datasets, disparity peaks occur much at much older ages when time-slicing rather than time binning approaches are used (Fig. \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
This is also true for stratigraphic time bins in the \cite{wright2017bayesian} dataset, although when using equal time bins the peaks are later than the time-slicing methods, or very similar (Fig. \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
Across the three time binning methods, the \cite{brusatte2014gradual} dataset has similar disparity peaks whichever method is used, the \cite{wright2017bayesian} dataset only had variation in peaks when using unequal time bins (stratigraphy), whereas in the \cite{bapst2016topology} and \cite{beckancient2014} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Fig. \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
Additionally, there seems to be small discrepancies within the time-slicing methods (gradual \textit{vs.} proximity) except in the \cite{beckancient2014} dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig. \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4)

@@

Of course the method also has limitations.
The main one of these is a practical one; it requires a time-calibrated phylogeny which are not available for all palaeontological datasets.
Furthermore, like most phylogeny based methods, time-slicing depends on ancestral state estimations.
Care should be taken in interpreting these, as they are highly dependent on the data and models used for the estimations \citep{Slateretal2012,doi:10.1080/10635150801910451}.
The difference between the time-binning and time-slicing methods could also simply be favoured towards time binning due to the nature of the fossil record itself.
The rates of sedimentation varies in time a space and can be really sporadic, influencing the groups found within and their temporal distribution (e.g. vertebrates, c.f. foraminifera).
In this case, different beds could represent different ``packages'' of fauna through time separated by gaps thus effectively binning the data rather than slicing.
For example, slicing through such strata will yield similar results whether where in time does the slice occur.
It is important to note however, that, as specified above, the time slicing method also includes ancestral estimations data (whether through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.
Finally, \cite{Hunt21042015} found that time series are best characterized by gradual directional changes (biased random walks).
In fact, homogeneous directional changes are more likely to be supported than heterogeneous ones (e.g. punctuated changes) in longer duration series with few samples per series.
In our implementation of time-slicing, the models are not selected based on any fit value (e.g. AIC) but merely on researcher's assumptions.
We thus suggest that both types of models (punctuated and gradual) are tested during analysis unless there could be strong independent support for one or the other.

@@

\paragraph{Acknowledgments.}
  NC thanks Mark Sutton and Philip Mannion for the invitation to contribute to the `Evolutionary Modelling' symposium at The Palaeontological Association Annual Meeting 2017.
  TG acknowledges support from the Australian Discovery Project Grant number DP170103227 awarded to Vera Weisbecker.
  We thank Dave Bapst, Graeme Lloyd, April Wright and David Wright for assistance in gathering data for the analyses and/or discussions about the approach; and Steve Brusatte, Sally Thomas and one anonymous reviewer for helpful comments on the manuscript.
